Higher-Order unification for free!

### 2024-04-09 paper/main.tex
typesetting

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different. For example:\\
%
\vspace{4pt}
{
\setlength{\tabcolsep}{2pt}
\begin{tabular}{lcl|}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
\end{tabular}
\phantom{mn}
\begin{tabular}{lcl}
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}
}
\vspace{4pt}
%
\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

\cref{sec:problem-statement} formally states the problem and gives the
intuition behind our solution. \cref{sec:encodings} discusses alternative
term encodings and related works. \cref{sec:grounwork} introduces
the languages \Fo{} and \Ho{}, \cref{sec:simulation} describes a
basic simulation of higher order logic programs.
\cref{sec:eta,sec:invariant1} completes its equational theory
with support for $\eta$-conversion. \cref{sec:beta} deals with the
practical necessity of ``tolerating'' terms outside of the
pattern-fragment and discusses how heuristic can be applied.
Finally \cref{sec:implementation} discusses the implementation in Elpi.

@@

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first class terms, i.e. they have a concrete syntax.
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}

\begin{figure}[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
\noindent
Unification variables (\elpiIn{fuva} term constructor)
in \Fo{} have no explicit scope:
the arguments of an higher order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp[fuva N, x]}>>, where \elpiIn{N} is a memory address and
\elpiIn{x} is a bound variable.\\
In \Ho the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher order and come equipped with an
explicit scope.
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

We simulate each run in \Fo{} with a run in \Ho as follows.
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \foUnifPb, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

\begin{definition}[\maybeeta]\label{def:maybeeta}
  $\maybeeta = \{ t ~|~ \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
\end{definition}

\noindent
An example of term $t$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda a.\lambda b.fba\}$
makes $\rho t = \lambda x.\lambda y.f x y$
that is the eta long form of $f$. This term is problematic since
its rigid part, the $\lambda$-abstractions, cannot justify a
unification failure against a constant or an application.

\begin{definition}[\maybebeta]\label{def:maybebeta}
  $\maybebeta = \{ X \appsep x_1 \ldots x_n ~|~ x_1 \ldots x_n ~\mathrm{are~distinct~names} \}$
\end{definition}

@@

\begin{definition}[\notllambda]\label{def:notllambda}
  $\notllambda = \{ X \appsep t_1 \ldots t_n ~|~ t_1 \ldots t_n ~\mathrm{are~not~distinct~names} \}$.
\end{definition}

@@

\noindent
\cref{prop:nf} is key to prove \cref{prop:simulation,prop:fidelity}:
the problematic terms moved on the side by the compiler are not
put back by \hstep, hence \Ue{} can operate properly.
In \cref{sec:compilation,sec:eta,sec:beta}
we describe how the compiler recognizes terms in \maybeeta, \maybebeta and
\notllambda and how progress takes care of them preserving \wellb
and granting \cref{prop:simulation,prop:fidelity}.

Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
since decompilation can introduce (actually restore) problematic terms.

@@

In this section we describe a basic compilation scheme that we refine
later, in the following sections. This scheme is sufficient to implement
an \Uo{} that respects $\beta$-conversion for terms in \llambda. The extension to
$\eta\beta$-conversion is described in Section \ref{sec:eta} and the support
for terms outside \llambda in Section \ref{sec:beta}.

>>>>

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different.
% For example:\\
% %
% \vspace{4pt}
% {
% \setlength{\tabcolsep}{2pt}
% \begin{tabular}{lcl|}
% \elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
% \elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
% \elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
% \end{tabular}
% \phantom{mn}
% \begin{tabular}{lcl}
% \elpiIn{P x} & \Ue{} & \elpiIn{x}\\
% \elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
% \elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
% \end{tabular}
% }
% \vspace{4pt}
%
\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

\todo{fix}\cref{sec:problem-statement} formally states the problem and gives the
intuition behind our solution. \cref{sec:encodings} discusses alternative
term encodings and related works. \cref{sec:grounwork} introduces
the languages \Fo{} and \Ho{}, \cref{sec:simulation} describes a
basic simulation of higher order logic programs.
\cref{sec:eta,sec:invariant1} completes its equational theory
with support for $\eta$-conversion. \cref{sec:beta} deals with the
practical necessity of ``tolerating'' terms outside of the
pattern-fragment and discusses how heuristic can be applied.
Finally \cref{sec:implementation} discusses the implementation in Elpi.

@@

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first class terms, i.e. they have a concrete syntax as
per \cref{code:common-terms}.
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure} %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
Unification variables (\elpiIn{fuva} term constructor)
in \Fo{} have no explicit scope:
the arguments of an higher order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp[fuva N, x]}>>, where \elpiIn{N} is a memory address and
\elpiIn{x} is a bound variable.\\
In \Ho the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher order and come equipped with an
explicit scope.
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

We simulate each run in \Fo{} with a run in \Ho as follows.
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\end{array}
$$
$$
\begin{array}{l}
  \runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \foUnifPb, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

\begin{definition}[\maybebeta]\label{def:maybebeta}
  \maybebeta is the set of terms of the form $X \appsep x_1 \ldots x_n$
  such that $x_1 \ldots x_n$ are distinct names (of bound variables).
\end{definition}

@@

\begin{definition}[\maybeeta]\label{def:maybeeta}
  \maybeeta is the set of terms $t$ such that $\exists \rho, \rho t$
  is an eta expansion.
\end{definition}

\noindent
An example of term $t$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$
makes $\rho t = \lambda x.\lambda y.f \appsep x \appsep y$
that is the eta long form of $f$. This term is problematic since
its leading $\lambda$ abstraction cannot justify a
unification failure against a constant or an application.

\begin{definition}[\notllambda]\label{def:notllambda}
  \notllambda is the set of terms of the form $X \appsep t_1 \ldots t_n$
  such that $t_1 \ldots t_n$ are not distinct names.
\end{definition}

@@

\noindent
Proposition \ref{prop:nf} is key to prove \cref{prop:simulation,prop:fidelity}:
informally it says that the problematic terms moved on the side by the compiler
are not put back by \hstep, hence \Ue{} can operate properly.
In \cref{sec:compilation,sec:eta,sec:beta}
we describe how the compiler recognizes terms in \maybebeta, \maybeeta and
\notllambda and how \progress takes care of them preserving \wellb
and granting \cref{prop:simulation,prop:fidelity}.

% Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
% since decompilation can introduce (actually restore) problematic terms.

@@

In this section we describe a basic compilation scheme that we refine
later, in the following sections. This scheme is sufficient to implement
a \hstep that respects $\beta$-conversion for terms in \llambda.
The extension to $\eta\beta$-conversion is described in \cref{sec:eta} and
the support for terms outside \llambda in \cref{sec:beta}.
